Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2024-01-05 doc/JAIR/lean-section.tex
Split Lean section as separate file

\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{booktabs}
\usepackage{hyperref}

\newcommand{\pand}{\mathbin{\land^\textsf{p}}}
\newcommand{\por}{\mathbin{\lor^\textsf{p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^\textsf{p}}
\DeclareMathOperator*{\Por}{\bigvee^\textsf{p}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}
\newcommand{\entail}{\vDash}
\newcommand{\entaildrat}{\entail_{\textrm{DRAT}}}

\newcommand{\assign}{\alpha}
\newcommand{\passign}{\rho}
\newcommand{\lassign}{\beta}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}
\newcommand{\validate}{\textsf{validate}}
\newcommand{\prov}{\textrm{Prov}}
\newcommand{\inputformula}{\phi_I}
\newcommand{\pogformula}{\theta_P}

\newcommand{\makenode}[1]{\mathbf{#1}}
\newcommand{\nodeu}{\makenode{u}}
\newcommand{\nodev}{\makenode{v}}
\newcommand{\nodes}{\makenode{s}}
\newcommand{\nodep}{\makenode{p}}
\newcommand{\noder}{\makenode{r}}

\newcommand{\simplify}[2]{#1|_{#2}}

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\cdfour}{\progname{CD4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}
\newcommand{\Dtrim}{\progname{Drat-trim}}

\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}

% Lean code:
\usepackage{listings}
%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
\def\lstlanguagefiles{lstlean.tex}
% set default language
\lstset{language=lean, xleftmargin=1em}
\lstset{backgroundcolor=\color{white}}

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}

\titlerunning{Certified Knowledge Compilation}

\author{Randal E. Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{Randy.Bryant@cs.cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrocki@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{Hoskinson Center for Formal Mathematics}
\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{Hoskinson Center for Formal Mathematics}
\author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}

\authorrunning{R. E. Bryant et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Randal E. Bryant, Wojciech Nawrocki, Jeremy Avigad, and Marijn J. H. Heule} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10003794</concept_id>
       <concept_desc>Theory of computation~Automated reasoning</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Automated reasoning}

%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
\keywords{Propositional model counting, Proof checking}

%\category{} %optional, e.g. invited paper

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\nolinenumbers %uncomment to disable line numbering

\EventEditors{Meena Mahajan and Friedrich Slivovsky}
\EventNoEds{2}
\EventLongTitle{26th International Conference on Theory and Applications of Satisfiability Testing (SAT 2023)}
\EventShortTitle{SAT 2023}
\EventAcronym{SAT}
\EventYear{2023}
\EventDate{July 4--8, 2023}
\EventLocation{Alghero, Italy}
\EventLogo{}
\SeriesVolume{271}
\ArticleNo{6}

\supplementdetails[subcategory=Paper Supplement]{DataPaper}{https://doi.org/10.5281/zenodo.7966174}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Access}
%\EventNoEds{2}
%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
%\EventShortTitle{CVIT 2016}
%\EventAcronym{CVIT}
%\EventYear{2016}
%\EventDate{December 24--27, 2016}
%\EventLocation{Little Whinging, United Kingdom}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\program}[1]{\textsc{#1}}
\newcommand{\lean}{Lean~4}

\begin{document}

\maketitle

\begin{abstract}

Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
is intractable on general representations. It can become tractable when formulas are expressed in a
special form, such as the decision-decomposable, negation normal form (dec-DNNF)\@.
\emph{Knowledge compilation} is the process of converting a formula
into such a form. Unfortunately existing knowledge compilers provide no guarantee that their output correctly
represents the original formula, and therefore they cannot validate a model count, or any other computed value.

We present \emph{Partitioned-Operation Graphs} (POGs), a form that can
encode all
of the representations used by existing knowledge compilers.
We have designed  CPOG, a framework that can express proofs of equivalence between a
POG  and a Boolean formula in conjunctive normal form (CNF).

We have developed a program that generates POG representations from dec-DNNF
graphs
produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

\section{Introduction}

Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
find an assignment satisfying $\phi$ or generate a proof that no
such assignment exists.  They have applications across a variety of
domains including computational mathematics, combinatorial
optimization, and the formal verification of hardware, software, and
security protocols.  Some applications, however, require going
beyond Boolean satisfiability.  For example, the \emph{model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

Model counting is a challenging problem---more challenging than the
already NP-hard Boolean satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal is to count models and not just determine
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters that scale to very large formulas have been developed, as
witnessed by the progress in recent model counting competitions.

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
(det-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
represent a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

One shortcoming of existing knowledge compilers is that they have no
generally accepted
way to validate that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that being able to generate proofs allow SAT solver developers to quickly
detect and diagnose bugs in their programs. This, in turn, has led
to more reliable SAT solvers.

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a form that can encode all of the representations produced by current knowledge
compilers. The CPOG (for ``certified'' POG) file format then
captures both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF).  A CPOG
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that, when satisified by a CPOG file, guarantees that it
encodes a well-formed POG and provides a valid equivalence proof.

\begin{figure}
\centering{
\begin{tikzpicture}[scale=0.17]
  \draw (8,10) rectangle (20,18);
  \node at (14,15.2) {Knowledge};
  \node at (14,12.8) {Compiler};

  \draw (28,10) rectangle (40,18);
  \node at (34,15.2) {Proof};
  \node at (34,12.8) {Generator};

  \draw (50,10) rectangle (62,18);
  \node at (56,15.2) {Proof};
  \node at (56,12.8) {Checker};

  \draw (50,0) rectangle (62,8);
  \node at (56,5.2) {Ring};
  \node at (56,2.8) {Evaluator};

  \draw[dashed] (48,-2) rectangle (64,22);
  \node at (56,20) {Trusted Code};

  %% CNF
  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
  \node [left] at (-1,20) {\large {$\inputformula$}};
  \node [above] at (2,20) {\texttt{.cnf}};

  %% NNF
  \draw[thick] (20,12) [-latex] -- (28,12);
  \node [above] at (24,12) {\texttt{.ddnnf}};

  %% CPOG
  \draw[thick] (40,12) [-latex] -- (50,12);
  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
  \node [above] at (44,12) {\texttt{.cpog}};

  %% OK/Not
  \draw[thick] (62,14) [-latex] -- (66,14) ;
  \node [right] at (66,15.2) { OK /} ;
  \node [right] at (66,12.8) { Not OK} ;
  \draw[thick] (62,4) [-latex] -- (66,4) ;
  \node [right] at (66,4) {\large {$\textbf{R}(\inputformula)$}};

\end{tikzpicture}
}
\caption{Certifying toolchain.
  The output of a standard knowledge compiler is converted into a combined graph/proof (CPOG)
  which can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a dec-DNNF representation,
and the \emph{proof
  generator} uses this to generate a CPOG file.
The \emph{proof checker} verifies the equivalence of the CNF and CPOG representations.
The \emph{ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

To ensure soundness of the abstract CPOG proof system, as well as
correctness of its concrete implementation, we formally verified the
proof system as well as versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a  CPOG file gives strong
assurance that the proof and the model count are correct. Our
experience with developing a formally verified proof checker has shown
that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of
requirements that guarantee soundness.  In fact, our efforts to
formally verify our proof framework exposed subtle conditions that we had
to impose on our partitioned sum rule.

We evaluate our toolchain using benchmark formulas from the 2022
standard and weighted model competitions.  Our tools handle all but
the largest dec-DNNF graphs generated by \dfour{}.  We also measure
the benefits of several optimizations as well as the relative
performance of the verified checker with one designed for high
performance and capacity.

We also show that our tools can provide end-to-end verification of
formulas that have been transformed by an equivalence-preserving
preprocessor.  That is, verification is based on the original formula,
and so proof checking certifies correct operation of the preprocessor,
the knowledge compiler, and the proof generator.

We have developed an online supplement to this
paper~\cite{bryant:sat:2023:supplement} that includes
a worked example, more details on the algorithms,
and extensive experimental results.

\section{Related Work}

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{heule:cade:2013,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
Our proofs construct a new propositional formula, and we must verify
%both
that it
%satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{section:formally-verified-toolchain}.

Capelli, Lagniez, and Marquis developed a knowledge compiler that
generates a certificate in a proof system that is itself based on
dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{}
program, a modified version of \dfour{}, generates annotations to the
compiled representation, providing information about how the compiled
version relates to the input clauses.  It also generates a file of
clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
Completing the certification involves running two different checkers
on the annotated dec-DNNF graph and the DRAT file.  Although the
authors make informal arguments regarding the soundness of their
frameworks, these do not provide strong levels of assurance.  Indeed,
we have identified a weakness in their methodology due to an
invalid assumption about the guarantees provided by \dtrim{}, the
program it uses to check the DRAT file.  This weakness is
\emph{exploitable}: their framework can be ``spoofed'' into accepting
an incorrect compilation.

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose during a top-down processing of
the input clauses.  Given input formula $\phi_I$, their first task is to
check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate program to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it does not necessarily
preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however, the
second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.

This weakness would allow a buggy (or malicious) version of \cdfour{}
to spoof the checking framework.  Given formula $\phi_1$ as input, it
could produce a compiled result, including annotations, based on $\phi_2$ and also include the second clause of $\phi_2$ in $R$.
The check with \dtrim{} would pass, as would the other tests performed
by their checker.  We have confirmed this possibility with
their compiler and checker.\footnote{Downloaded May 18, 2023 as\\
\url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}

This weakness can be corrected by restricting \dtrim{} to only add
clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
command-line argument to \dtrim{} that enforces this
restriction.\footnote{Available at
\url{https://github.com/marijnheule/drat-trim/releases/tag/v05.22.2023}.}  This weakness, however, illustrates the general challenge of
developing a new proof framework.
As we can attest,
without engaging in an effort to formally verify the framework, there are likely to be
conditions that make the framework unsound.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithms commonly used by model counters.  They
developed a program that can generate proof traces from dec-DNNF
graphs and a program to check adherence to their proof rules.  This
framework is not directly comparable to ours, since it only certifies
the unweighted model count, but it has similar goals.
Again, they provide only  informal arguments
regarding the soundness of their framework.

Both of these prior certification frameworks are strongly tied to the
algorithms used by the knowledge compilers and model counters.  Some
of the conditions to be checked are relevant only to specific
implementations.    Our framework is very general and is based on a small set
of proof rules.  It builds on the highly developed
concepts of clausal proof systems.  These factors were important in enabling formal verification.
In Section~\ref{sect:experimental} and the supplement~\cite{bryant:sat:2023:supplement},
we also compare the performance of our toolchain to these other two.

\section{Logical Foundations}
\label{section:logical:foundations}

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an \emph{assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
  \subseteq \varset$.  We say the assignment is \emph{total} when it
  assigns a value to every variable ($\varset' = \varset$), and that
  it is \emph{partial} otherwise.
  The set of all possible total assignments over
  $\varset$ is denoted $\uassign$.

For each variable $x \in \varset$,
  we define the \emph{literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated \emph{dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of \emph{models} $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight some special classes of Boolean formulas.  A formula is
  in \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
  A \emph{partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{definition}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

  \section{Ring Evaluation of a Boolean Formula}

We propose a general framework for summarizing properties of Boolean
formulas along the lines of algebraic model counting~\cite{kimmig:jal:2017}.

\begin{definition}[Commutative Ring]
  A \emph{commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{definition}
We write $a - b$ as a shorthand for $a + -b$.

\begin{definition}[Ring Evaluation Problem]
\label{def:ring_evaluation}
  For commutative ring $\ring$, a \emph{ring weight function} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.

  For Boolean formula $\phi$ and ring weight function $w$, the \emph{ring evaluation problem} computes
  \begin{equation}
    \begin{array}{rcl}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
    \end{array}
  \end{equation}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:weight}
\end{definition}

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
(standard) \emph{model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the \emph{density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

The \emph{weighted model counting}  problem is also defined over
rational numbers.  Some formulations  allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The \emph{function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, choose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent,
and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
For example, it can compare different runs on a single formula,
either from different compilers or from a single compiler with different configuration parameters.

%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.

\section{Partitioned-Operation Graphs (POGs)}
\label{sect:pog}

>>>>

%% Separate file for Lean presentation

\section{A Formally Verified Toolchain}
\label{sect:formally-verified-toolchain}
\label{sect:lean:subtle-condition}

We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}). These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}. Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic. In this section, we briefly describe the functionality we implemented and what we proved about it.

\vspace{1em}\noindent
\textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$. The POG being constructed, and the set of active clauses are stored in the checker state \texttt{st}. The checker begins by parsing the input formula, initializing the active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one. It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect. Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}. Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\ref{thm:lean:equiv}).

The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{sect:logical:foundations}. Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas. \lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}. Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}. Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$. We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
